Distributional Trees and Forests

### 2017-11-30 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1783 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b2edc10747e23a82ab71433996ebfedfbd688f12

\subsection{Distributional Forest}
\label{distforest}
A distributional forest consists of an ensemble of distributional trees $\{t_s\}_{s=1,\ldots,T}$. Each of these trees is built on a different data set which can either be a bootstrap or a subset of the original data set. Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables. In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\citealt{breiman2001random}).\\
\\
Forest models differ mainly in the way how the trees are combined to make predictions. In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.\\
Contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new set of covariates. This difference is represented in the weight vector as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights \\
$\bold{\omega}^{forest} = (\omega^{forest}_1, \ldots, \omega^{forest}_n)$.\\
For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each tuple $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$. 
$$
\omega^{forest}_i(\bold{z}^{\ast}) = \sum_{s=1}^T \sum_{b=1}^{B^s} I((\bold{z}_i \in \mathcal{B}^s_b) \land (\bold{z}^{\ast} \in \mathcal{B}^s_b))
$$
where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.\\
This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new set of covariates $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
$$
\hat{\theta}(\bold{z}^{\ast}) =  \max_{\theta \in \Theta} \sum_{i=1}^n \omega^{forest}_i(\bold{z}^{\ast}) \cdot \ell(\theta; y_i)
$$
While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates due to the individual weights calculated separately for each set of covariates.\\
\\
As this novel methodology combines the advantages of smooth and flexible parametric modeling with those of tree structured modeling it provides a model that can deal with a wide range of situations. Abrupt changes and non-additive effects can be captured by its tree structure while smooth effects can be modeled reasonably well by combining various tree and smoothing the steps imposed by a single tree. As scores are considered for each distribution parameter (separate) effects in different parameters can be captured. At the same time, also interactions are detected automatically by splitting the data in a way that is sensitive to changes in either each of the parameters or combined in various parameters.\\
Even though many situations demand for the use of specificaly taylored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required. In these cases a distributional forest is a recommendable compromise providing reasonable models and performing well without any prespecification or additional knowledge needed in advance.

@@

For the prespecified GAM model the covariates were selected manually based on meteorological knowledge and experience. Additionally, interactions are specified in advance, once between \texttt{tp{\_}mean1218} and \texttt{cape{\_}mean1218} for the location parameter and once between \texttt{tp{\_}sprd1218} and \texttt{cape{\_}mean1218} for the scale parameter. These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation. By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information and is therefore expected to profit from this advantage.\\
In the EMOS model the logarithmic function is chosen as the link function for the scale parameter. This choice has been made based on the results of \citealt{Gebetsberger2017} where different link functions (identity, quadratic and logarithmic function) have been compared. Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function lead to the best performance of the EMOS model in this application.\\
\\
In order to compare the performance of the chosen methods, the continuous ranked probability score (CRPS) is calculated for each model. The CRPS measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
$$
\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
$$
where \(\mathbf{1}\) is the indicator function. \\
In this application the CRPS is evaluated out of sample on a data set that has not been used to build the model. The mean CRPS over thi data set is used for validation. Since the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit. In applications, one big advantage in terms of interpretability is that the resulting value is on the scale of the observation. \\
For the purpose of comparing the performance of a distributional forest to those of other models the so-called skills score can be calculated with respect to the CRPS and one model being chosen as a reference. In that way we obtain a measure of how much a method improves the model performance in terms of CRPS compared to the reference method.
$$
\text{SkillsScore}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
$$

@@

Figure~\ref{map} shows the location of all stations in the data set. Each observation station is illustrated by a symbol with the type of symbol representing the method that performed best in terms of CRPS at this station. The color of the symbol indicates the difference of the CRPS of the forest model and the CRPS of the best perfoming among the other three models (prespecified GAM, boosted GAM and EMOS).

>>>>

\subsection{Distributional Forest}
\label{distforest}
A distributional forest consists of an ensemble of distributional trees $\{t_s\}_{s=1,\ldots,T}$. Each of these trees is built on a different data set which can either be a bootstrap or a subset of the original data set. Moreover, in each node a subset of the partitioning variables is chosen randomly as candidates for splitting variables. In that way the correlation among the resulting trees is reduced and therefore also the variance of the model (\citealt{breiman2001random}).\\
\\
Forest models differ mainly in the way how the trees are combined to make predictions. In this case the forest itself is only used to evaluate neighborhood relationships among the observations which can then be included in the fitting process.\\
Contrary to distributional trees, in distributional forests the whole learning data set can be used to fit the distributional model and specify the predicted distribution for a new set of covariates. This difference is represented in the weight vector as the tree weights $\bold{\omega}^{tree}$ containing only zeros and ones are now replaced by integer valued forest weights \\
$\bold{\omega}^{forest} = (\omega^{forest}_1, \ldots, \omega^{forest}_n)$.\\
For a new set of covariates $\bold{z}^{\ast} = (z^{\ast}_1, \ldots, z^{\ast}_m) $ the set of weights is obtained by the following procedure: For each tuple $(y_j,\bold{z}_j)$ in the learning data set $\{(y_i,\bold{z}_i)\}_{i=1\ldots,n}$ a weight is calculated by counting the number of trees in which $(y_j,\bold{z}_j)$ ends up in the same terminal node as the new set of covariates $\bold{z}^{\ast}$. 
$$
\omega^{forest}_i(\bold{z}^{\ast}) = \sum_{s=1}^T \sum_{b=1}^{B^s} I((\bold{z}_i \in \mathcal{B}^s_b) \land (\bold{z}^{\ast} \in \mathcal{B}^s_b))
$$
where $T$ is the number of trees of the forest, $B^s$ is the number of terminal nodes of the \textit{s}-th tree and $\mathcal{B}^s_b$ represents the \textit{b}-th terminal node of the \textit{s}-th tree.\\
This individual set of weights $\bold{\omega}^{forest}(\bold{z}^{\ast}) = (\omega^{forest}_1(\bold{z}^{\ast}), \ldots, \omega^{forest}_n(\bold{z}^{\ast}))$ for the new set of covariates $\bold{z}^{\ast}$ can now be included in the estimation process of the distribution parameters and in that way leads to its individual parameter vector $\hat{\theta}(\bold{z}^{\ast})$.
$$
\hat{\theta}(\bold{z}^{\ast}) =  \max_{\theta \in \Theta} \sum_{i=1}^n \omega^{forest}_i(\bold{z}^{\ast}) \cdot \ell(\theta; y_i)
$$
While in a distributional tree the same distribution is specified for all sets of covariates being in the same node, a distributional forest predicts and fits a different distribution for each different set of covariates due to the individual weights calculated separately for each set of covariates.\\
\\
As this novel methodology combines the advantages of smooth and flexible parametric modeling with those of tree structured modeling it provides a model that can deal with a wide range of situations. Abrupt changes and non-additive effects can be captured by its tree structure while smooth effects can be modeled reasonably well by combining various tree and smoothing the steps imposed by a single tree. As scores are considered for each distribution parameter (separate) effects in different parameters can be captured. At the same time, also interactions are detected automatically by splitting the data in a way that is sensitive to changes in either each of the parameters or combined in various parameters.\\
Even though many situations demand for the use of specifically tailored models providing the needed features and therefore being clearly advantageous and the best choice, it is often hard to tell in advance which features will be required. In these cases a distributional forest is a recommendable compromise providing reasonable models and performing well without any prespecification or additional knowledge needed in advance.

@@

For the prespecified GAM model the covariates were selected manually based on meteorological knowledge and experience. Additionally, interactions are specified in advance, once between \texttt{tp{\_}mean1218} and \texttt{cape{\_}mean1218} for the location parameter and once between \texttt{tp{\_}sprd1218} and \texttt{cape{\_}mean1218} for the scale parameter. These interactions might capture events with a high potential for moderate or severe thunderstorms with large amounts of precipitation. By putting extra effort in the specification of the formulas the prespecified GAM model is provided with additional information and is therefore expected to profit from this advantage.\\
In the EMOS model the logarithmic function is chosen as the link function for the scale parameter. This choice has been made based on the results of \citealt{Gebetsberger2017} where different link functions (identity, quadratic and logarithmic function) have been compared. Moreover, testing the identity, quadratic and logarithmic function as link functions in this setting supported this choice as the logarithmic function lead to the best performance of the EMOS model in this application.\\
\\
In order to compare the performance of the chosen methods, the continuous ranked probability score (CRPS) is calculated for each model. The CRPS measures the accuracy of a probabilistic model. For one observation $y$ with the predicted distribution function $F$ the CRPS value is calculated by
$$
\text{CRPS}(y, F) = \int_{-\infty}^\infty (F(z) - \mathbf{1}(y \leq z))^2 dz
$$
where \(\mathbf{1}\) is the indicator function. \\
In this application the CRPS is evaluated out of sample on a data set that has not been used to build the model. The mean CRPS over this data set is used for validation. Since the CRPS can be seen as a generalization of the mean absolute error, a low value indicates a good fit. In applications, one big advantage in terms of interpretability is that the resulting value is on the scale of the observation. \\
For the purpose of comparing the performance of a distributional forest to those of other models the so-called skills score can be calculated with respect to the CRPS and one model being chosen as a reference. In that way we obtain a measure of how much a method improves the model performance in terms of CRPS compared to the reference method.
$$
\text{SkillsScore}_{\text{method}} = 1 - \frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{reference}}}
$$

@@

Figure~\ref{map} shows the location of all stations in the data set. Each observation station is illustrated by a symbol with the type of symbol representing the method that performed best in terms of CRPS at this station. The color of the symbol indicates the difference of the CRPS of the forest model and the CRPS of the best performing among the other three models (prespecified GAM, boosted GAM and EMOS).
